Changes in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2019-12-20 analysis/paper/paper.Rmd
working on intro and background

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it it thought to reflect more specialed hunting practices (citations). Stemmed points are considered to be projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using behavioural ecology to investigate changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations). To our knowledge, the implications for forager mobility and site use have not been explored. Most Korean Paleolithic sites do not preserve organic remains to understand forager subsistence economies, so we lack direct faunal evidence on hunting strategies. We attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmntal changes.

<!--
Need something about connection between stemmed point and specifics of hunting, what kind delivery system? What kind of game?
-->

@@

## Theory and background to our three questions

<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q

State the questions very clearly and give background into relevant literature:

 -  the population: that some relationship may exist between population as a cause of techno change: Torrence, Collard, Shennan

 - the way people used the landscape: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott

 - the way people used sites: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock

We ask two questions to invetigate the period around the introduction of stemmed points. First, we are asking how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments? The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating foraging strategies. In addition, the new tool production technique required hunter-gatherers to move for acquiring specific raw materials [@blumenschine2008effects; @reti2016quantifying]. According to ethnographic observations of mobility, people move frequently with little knowledge of the places for foraging and acquiring raw materials purpose with portable, mulifunctional, long lasting tools[@clarkson2008lithics;@kuhn2014mousterian]. The study reported here was designed to interpret the impact of the new tools on the landscape use during the transitional period.

<!-- I think we were also talking about example that a new tech actually changed the landscape. But I'm not sure where I can put the example

For example, the new technology could change their landscape. Hominin fire use in the early Pleistocene has been studied and some argue that usage of fire changed the landscape and their dietary [@pruetz2010brief; @glikson2013fire; @parker2016pyrophilic; @hlubik2019hominin].

Second, we focus on how the new technology changed the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp. The stone artifact assemblage from a site can inform us how long individuals or groupd stayed or what they did there so we can assume the way people use the site [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. Stone artifact contains a record of material transport and selection because it was carefully designed so that people could make a living from the landscape that constraints the production, efficiency, use, transportation, and maintenance [@clarkson2008lithics]. Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact]. This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period and the relationship with the appearance of new tools.

Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside. The demands derived from significan population dynamic in addition to environmental change could lead to development of the tools [buchanan2016drivers; @mihailovic2016technological; prufer2019linking]. Some studies argue that technological complexity and richness are highly influenced by population size because in large population, because new innovation could occur during the process of learning from others [@shennan2001demography; @henrich2004demography; @collared2005causes; @collard2013risk; @buchanan2016drivers]. Therefore, we will check the population change during the transitional period to understand its impact on the technology.

-------------- previous paragraphs ------

@@

## Regional Archaeology

Northeast Asia/similar studies and previous research in the adjacent regions : General description of EUP in East Asia

The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic. In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois [@gao2002critique; @Bar-Yosef_2002; @mellars2007rethinking; @norton2009evolution; @norton2009east; @lee2013current]]. Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].

In China, the transition from the Middle to the Upper Paleolithic dates to 35-30 ka appeared with a few body decorations, bone tools, core-and-flake in North and cobble tools in South. In addition to the technological changes, the use of symbols in the body decorations reflects social and cultural changes occurred during the onset of the Chinese Upper Paleolithic [@qu2013chinese; @peng2018onset]. At the Shuidonggou site in northern China which is considered as a typical site for showing the transition from cobble to blade tool industries in Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou], newly discovered localities (SDG 2, 7-12) show various aspects of the initial Late Paleolithic from 32,000 BP to 6000 BP. Some sites (SDG 9, 12) have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12), while other sites have mostly small, irregular flakes [@Gao_Zhang_Yang_Shen_Wu_2010]. The transition to the Upper Paleolihtic in Japan is under the long term debate due to the uncertainty about the existence of the previous period in the Archipelago. But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new foraging strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. The technological transition in Korea is mainly characterized by the early appearance of stemmed points and following blade technology. Using the blade as a blank, the stemmed points became more standardized and increased in number. The flake tools became more sophisticated, with a greater diversity in both type and size along with continuously used core tools during the Upper Paleolithic [@Bae_2010; @Lee_2013; @Lee_2016].

In this research, we choose to focus on the technological transition and its impact on mobility and site occupation pattern in Korea because of the uniqueness of sudden and earliest appearance of stemmed points among the broad area of Northeast Asia during the Late Pleistocene. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. The stemmed point is a projectile point made out of elongated flake or blade and with gentle retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. Combining with the blade industry, the shape became more standardized with one or two ridges on dorsal side and triangular horizontal cross section. The stemmed point is often called as a tanged point, but we prefer to use stemmed point to distinguish with other stone projectile points of Bronze age in Korea, also called as a tanged point in English literature. The early appearance of the stemmed points in the Korean Peninsula brings about several issues to consider such as issues of human migration or advanced cognitive ability to make a new tool adapting to the surrounding environment. As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal [@Keeley_1982; @kuhn2015artifacts]. The new projectile point also indicates a new way of hunting methods or other foraging strategies because the throwing technique enables to earn a long-distance target. In addition, this new tool spread to adjacent regions such as Japanese Archipelago by combination of watercraft technique and social network so that the study of stemmed points play an important role in understanding about peopling in Japan and the relationship between the Peninsula and Archipelago. This project focuses on the functional aspect to understand the technology and foraging pattern based on the concept of HBE.

>>>>

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices (citations). Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations). To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- sgenerally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) -->

@@

### Regional archaeology

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals (citation).

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) (citation). The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain. <!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate? -->

The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools (citations). New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools. <!-- need some ages for these types of evidence --> These have been interpreted as evidence of new foraging strategies possibly driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. <!-- need to expand, mention some key site, blades appear when? what is the debate? -->

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. <!-- need to expand, mention some key site & details. what is the debate?  -->

### Behavioural ecology and forager land use behaviours

Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages. We focus on the appearance of stemmed points in Korea because these are one of the earliest new technologies that definiately represents a new hunting methods because the halfting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We ask two questions to invetigate the period around the introduction of stemmed points in Korea.

First, we ask how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments? The new technology might allow foragers to explore and stay in different or harsher territories by increasing mobility. <!-- need something about how technology and landscape use is linked ... - the way people used the *landscape*: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott  ... -->

Second, we focus on how the new technology changed the way people use their habitation sites. After the appearance of these new tools, did they tend to stay in one location longer, or for shorter periods, perhaps for specific purposes?. Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact]. The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insignts into what they did there [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. Stone artifacts can contain a record of raw material transport and selection and give insights into land-use behaviour [@clarkson2008lithics]. <!-- - the way people used *sites*: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock -->

<!-- Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside. The demands derived from significan population dynamic in addition to environmental change could lead to development of the tools [buchanan2016drivers; @mihailovic2016technological; prufer2019linking]. Some studies argue that technological complexity and richness are highly influenced by population size because in large population, because new innovation could occur during the process of learning from others [@shennan2001demography; @henrich2004demography; @collared2005causes; @collard2013risk; @buchanan2016drivers]. Therefore, we will check the population change during the transitional period to understand its impact on the technology. -->

<!--  -------------- previous paragraphs ------

@@
